---
title: "Figure 6"
author: "Monica Mugnier"
date: "7/12/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(igraph)
library(tidyverse)
library(reshape2)

```

#Generate Plots for Figure 6:
##Expressed VSG C-termini are primarily type 1 and type 2

C-terminal BLASTp-tree was made by Bill Wickstead

## Dataframes
Set up the Tbg VSGs and Cross et al -typed VSG dataframes including their e-values output from the blast (this df includes both Tb gambiense + TB427 VSGs).
```{r}
#upload output .csv file from blastdb
Cterm_links <- data.frame(read.csv("FigureData/CterminalNetwork.csv", header=F))
#View(Cterm_links)

#rename columns
Cterm_links <- Cterm_links %>%
  rename(from = V1,
         to = V2,
         evalue = V3)

```

```{r}
#upload Cross et al table, 545 VSGs with a C-terminal type (remove the ones called 'x', which were not actually typed)
Cterm_types <- read.csv("FigureData/Cross et al.csv", header=T, as.is=T) 
#rename columns
Cterm_types <- separate(Cterm_types, X.ID, sep = "-", into = c("Tb", "num"))
Cterm_types$seqs <- paste(Cterm_types$Tb, "VSG-", Cterm_types$num, sep="")
Cterm_types <- subset(Cterm_types, select = c("seqs", "Type"))
#The below is just adding a column of colors for each type. Once you get your own typing sytem I would recommend rewriting this into a more simple format to apply to your unique "nodes", but I'm leaving it like this for easy access for now.
Cterm_types$Colors <- ifelse(Cterm_types$Type =="2-2", "blue", Cterm_types$Type)
Cterm_types$Colors <- ifelse(Cterm_types$Type =="2-1", "red", Cterm_types$Colors)
Cterm_types$Colors <- ifelse(Cterm_types$Type =="1", "green", Cterm_types$Colors)
Cterm_types$Colors <- ifelse(Cterm_types$Type =="3", "black", Cterm_types$Colors)
Cterm_types$Colors <- ifelse(Cterm_types$Type =="5", "yellow", Cterm_types$Colors)
#Cterm_types$Colors <- ifelse(Cterm_types$Type =="x", "brown", Cterm_types$Colors)


#adding patient seqs to Cterm_types df and assigning them as type "patient"
pat_links <- arrange(unique(Cterm_links[1]), from)
#View(pat_links)
pat_links <- pat_links[grep("TRINITY", pat_links$from, value = F), ]
pat_types <- data.frame(pat_links)
#View(pat_types)
pat_types <- pat_types %>%
  rename(seqs = pat_links)
pat_types$Type <- "Patient"
pat_types$Colors <- "purple"


#adding Tbg seqs to Cterm_types df and assigning them as type "Tbg"
tbg_links <- arrange(unique(Cterm_links[1]), from)
#View(pat_links)
tbg_links <- tbg_links[grep("Tbg", tbg_links$from, value = F), ]
tbg_types <- data.frame(tbg_links)
#View(pat_types)
tbg_types <- tbg_types %>%
  rename(seqs = tbg_links)
tbg_types$Type <- "Tbg"
tbg_types$Colors <- "brown"

Cterm_types_m <- rbind(Cterm_types, pat_types, tbg_types) %>%
  filter(Type != 'x')
#View(Cterm_types_m)
```

```{r}
#keep only the VSGs that have an annotation
annotated_Cterm_links <- inner_join(Cterm_links, Cterm_types_m, by = c("to" = "seqs")) #This is comparing the column "to" in "Cterm_links" dataframe to the "seqs" column in "Cterm_types" dataframe and removing all the rows that are the same.
annotated_Cterm_links <- inner_join(annotated_Cterm_links, Cterm_types_m, by = c("from" = "seqs")) #This is doing the same thing as the previous dataframe, but comparing "from" in "excluded_Cterm_links" to "seqs" in "Ctermtypes"
#View(excluded_Cterm_links)
```

## Network plots
Visualize VSG homology with igraph - creates network plots and clusters each group of VSG C-termini.
```{r}
#Once you've excluded sequences you don't have typing for already, you want to create your "nodes" that igraph will need. I found the easiest way to do this was to remove any duplicate values in the "from" column.
Cterm_nodes <- arrange(unique(annotated_Cterm_links[1]), from)
#View(Cterm_nodes)


#be sure to use C-term_types file that has combined both Tb427 and patient data set
tester <- inner_join(Cterm_nodes, Cterm_types_m,  by = c("from" = "seqs")) #This will be your "nodes" input for igraph later. These are the vertices you'll see on the graph and you'll refer to the color in here to make the graph look presentable.
#View(tester)

Cterm_self <- annotated_Cterm_links[!(annotated_Cterm_links$from == annotated_Cterm_links$to),] #Removing sequences that map to themselves so shorten your analysis by a little.
#View(Cterm_self)
```

```{r}
#cutoff of 1e-4 creates a network plot which aligns to previous typing systems, also looked at cutoff 1e-5 to get more separation between clusters
#Different evalue cutoffs to generate side by side graphs showing the shape changes
Cterm_links_cutoff4 <- filter(Cterm_self, evalue < 1e-4)
Cterm_links_cutoff5 <- filter(Cterm_self, evalue < 1e-5)

#This is the actual plot build where you change the vertices and edges.
nets4 <- graph_from_data_frame(d=Cterm_links_cutoff4, vertices=tester, directed=F)
nets5 <- graph_from_data_frame(d=Cterm_links_cutoff5, vertices=tester, directed=F)

#Removing any residual loops where the vertex goes to itself and also removing multiple edges between the same vertices.
nets_simple4 <- igraph::simplify(nets4, remove.multiple = T, remove.loops = T)
nets_simple5 <- igraph::simplify(nets5, remove.multiple = T, remove.loops = T)

#The two sections below create new maps to delete vertices that are unused. It wouldn't be accurate to present data like this, but it is helpful to get an idea of how many points remain relevant as your evalues change.
Isolated.4 = which(degree(nets_simple4)==0)
Isolated.5 = which(degree(nets_simple5)==0)

nets_simple.4 = delete.vertices(nets_simple4, Isolated.4)
nets_simple.5 = delete.vertices(nets_simple5, Isolated.5)

cluster_eigen4 <- cluster_leading_eigen(nets_simple4, weights = NA)
cluster_eigen5 <- cluster_leading_eigen(nets_simple5, weights = NA)

#view which VSGs belong to which cluster
#membership(cluster_eigen4)
#communities(cluster_eigen4)
#membership(cluster_eigen5) 
#communities(cluster_eigen5) 
```

```{r}
#extract community information
group_ids <- lapply(tester %>% split(.$Type), function(grp) { grp$from })
#from Jaime:
# this is messy, but here's how to make a dataframe of membership so I can merge types
# extract membership from the eigen cluster list. 
# these are the numbers associated with each eigen cluster
community5 <- cluster_eigen5$membership
#community5

# extract seq ID from the network plot dataframe
v5 <- as_ids(V(nets_simple5))

# each of these can be the column of the dataframe
v5com <- data.frame(v5, community5)
#View(v5com)

#rename columns
colnames(v5com) <- c("seqs", "eigen_cluster")

#create table with seq# + eigen cluster community
v5com <- merge(v5com, Cterm_types_m)
v5com$eigen_cluster <- as.factor(v5com$eigen_cluster)

v5com$Type <- factor(v5com$Type, levels = c("1", "2-1", "2-2", "3", "5", "Patient", "Tbg"))

com5_ids <- lapply(v5com %>% split(.$eigen_cluster), function(grp) { grp$VSG })
#assign colors
color5 <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00")
v5com$Colors <- color5[v5com$eigen_cluster]

#give our color palette to the typed VSGs
SarahPalette <- c("#ffdec2", "#fcab64", "#f27618", "#91a6b4", "#02263d", "red")

color_palette <- c("#ED7D31", "#FFC000", "#FFD966", "#A9D18E", "#9DC3E6", "#8D2E94", "#9A132D")
#assign each type a color
v5com$type_colors <- ifelse(v5com$Type == 1, "#ED7D31", v5com$Type)
v5com$type_colors <- ifelse(v5com$Type == "2-1", "#FFC000", v5com$type_colors)
v5com$type_colors <- ifelse(v5com$Type == "2-2", "#FFD966", v5com$type_colors)
v5com$type_colors <- ifelse(v5com$Type ==3, "#A9D18E", v5com$type_colors)
v5com$type_colors <- ifelse(v5com$Type ==5, "#9DC3E6", v5com$type_colors)
v5com$type_colors <- ifelse(v5com$Type =="Patient", "#8D2E94", v5com$type_colors)
v5com$type_colors <- ifelse(v5com$Type =="Tbg", "#9A132D", v5com$type_colors)
```

##Figure 6B, C
```{r}
#create network plot
#create "Figures" folder within working directory

#1e-5 plot
#pdf("Figures/1e-5 evalue.pdf", width = 11, height = 8.5) #saving the figure. You already have to have the "Figures" folder in your working directory
par(mfrow=c(1,2))
set.seed(1234)
plot(cluster_eigen5 #cluster communities
     , nets_simple5 #maps the vertices and edges
     , edge.color = "grey50" 
     , vertex.size = 7
     , asp = 0
     , edge.arrow.size = 0
     , edge.width = .5
     , vertex.label = NA)
title("Eigen cluster communities", cex.main = 1) #plot title
legend("topleft", bty = "n",
       legend = levels(v5com$eigen_cluster),
       col= color5[as.numeric(levels(v5com$eigen_cluster))],pch = 19, pt.cex = 1, cex=1, horiz = FALSE)


set.seed(1234)
plot(nets_simple5
     , edge.color = "grey50" 
     , vertex.size = 7
     , asp = 0
     , edge.arrow.size = 0
     , edge.width = .5
     , vertex.label = NA
     , vertex.color = v5com$type_colors)
title("Cross typing", cex.main = 1)
legend('topleft', legend = names(group_ids),
       col = color_palette,
       pch = 19, bty = "n",  pt.cex = 1, cex = 1, 
       text.col = "black", horiz = FALSE)
```


## Summary Table
```{r}
v5com<-data.frame(v5com)
Tbg<-v5com %>%
  filter(Type == "Tbg")
#View(Tbg)

Tbg_type <-  mutate(Tbg,
         cross_type = ifelse(eigen_cluster == 1, "2-1/2-2",
                             ifelse(eigen_cluster == 2, "3",
                                    ifelse(eigen_cluster == 3, "1",
                                           ifelse(eigen_cluster == 4, "5",
                                                  ifelse(eigen_cluster == 5, "1", "oops")))))
  ) %>% 
  select(c("seqs", "eigen_cluster", "cross_type", "Type")) %>% 
  rename(sample = Type)



Patient <-v5com %>%
  filter(Type == "Patient")


Patient_type <-  mutate(Patient,
         cross_type = ifelse(eigen_cluster == 1, "2-1/2-2",
                             ifelse(eigen_cluster == 2, "3",
                                    ifelse(eigen_cluster == 3, "1",
                                           ifelse(eigen_cluster == 4, "5",
                                                  ifelse(eigen_cluster == 5, "1", "oops")))))) %>%
  select(c("seqs", "eigen_cluster", "cross_type", "Type")) %>% 
  rename(sample = Type)



Cross_type <- v5com %>% 
  filter(str_detect(seqs, "Tb427")) %>%
  rename(cross_type = Type) %>%
  select(c("seqs", "eigen_cluster", "cross_type")) %>% 
  mutate(sample = "Tb427")

Cross_type$cross_type <- as.character(Cross_type$cross_type)
Cross_type$cross_type <- ifelse(Cross_type$cross_type == "2-1", "2-1/2-2",
                               ifelse(Cross_type$cross_type == "2-2", "2-1/2-2",
                                      Cross_type$cross_type))

Cross_type$cross_type <- as.factor(Cross_type$cross_type)

#510 of the 545 typed VSGs which had a C-terminus detectable by HMMer.
#missing VSGs:
setdiff(Cterm_types$seqs[Cterm_types$Type != 'x'], Cross_type$seqs)
length(setdiff(Cterm_types$seqs[Cterm_types$Type != 'x'], Cross_type$seqs))
#VSG-22 is missing from George's databases. I'm not sure if this is meaningful but it will have to be excluded  

#View(Tbg_type)
#View(Patient_type)
#View(Cross_type)


```
## Analysis
```{r}

#combine patient + Tb427 + Tbg df into one including C-terminal type for all to compare across datasets

All_types <- rbind(Cross_type, Patient_type, Tbg_type)
All_types$eigen_cluster <- as.factor(All_types$eigen_cluster)


```

### Summary plots
```{r}
#summary graphs
#Tb427 vs patient VSG vs Tbg stacked bar by Cross type

df <- All_types %>%
  group_by(sample, cross_type) %>%
  summarise(count = n())
df

df$sample <- factor(df$sample, levels = c("Tb427", "Patient", "Tbg"))


#color-blind friendly palette: c("#ED7D31", "#FFC000", "#71A250", "#9DC3E6", "#966398") 
#color_palette <- c("#ED7D31", "#FFC000", "#FFD966", "#A9D18E", "#9DC3E6", "#8D2E94", "#9A132D")

color_palette2 <- c("#ED7D31", "#FFC000", "#A9D18E", "#9DC3E6")
#1, 2-1, 2-2, 3, 5, p, x
df %>%
  ggplot(aes(sample, count, fill = cross_type)) +
  geom_bar(position = "fill", stat = "identity", width=0.5)  +
  scale_fill_manual(values=color_palette2) +
  #theme(legend.position = "none") +
  labs(title = "", 
       x = "", 
       y = "VSG proportion",
       fill = "Cross type") + theme_classic(base_size = 22) +  
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, size = 20), axis.text.y = element_text(size = 20), axis.title = element_text(size = 20),legend.title = element_text(size = 16), legend.text = element_text(size = 14)) 
#  ggsave(path = "Figures", filename = "Tbg.pat.pdf", device = "pdf", width = 8, height = 6)

```

###Figure 6D
Based on Eigen Cluster identification - this is more appropriate, because typing for Tbg and patient VSGs is based on these clusters.
```{r}

All_types_eigen <- mutate(All_types,
         eigen_type = ifelse(eigen_cluster == 1, "2-1/2-2",
                             ifelse(eigen_cluster == 2, "3",
                                    ifelse(eigen_cluster == 3, "1",
                                           ifelse(eigen_cluster == 4, "5",
                                                  ifelse(eigen_cluster == 5, "1", "oops"))))))


df_eigen <- All_types_eigen %>%
  group_by(sample, eigen_type) %>%
  summarise(count = n())
df_eigen

df_eigen$sample <- factor(df_eigen$sample, levels = c("Tb427", "Patient", "Tbg"))


#color-blind friendly palette: c("#ED7D31", "#FFC000", "#71A250", "#9DC3E6", "#966398") 
#color_palette <- c("#ED7D31", "#FFC000", "#FFD966", "#A9D18E", "#9DC3E6", "#8D2E94", "#9A132D")

color_palette2 <- c("#ED7D31", "#FFC000", "#A9D18E", "#9DC3E6")
#1, 2-1, 2-2, 3, 5, p, x
df_eigen %>%
  ggplot(aes(sample, count, fill = eigen_type)) +
  geom_bar(position = "fill", stat = "identity", width=0.5)  +
  scale_fill_manual(values=color_palette2) +
  #theme(legend.position = "none") +
  labs(title = "", 
       x = "", 
       y = "VSG proportion",
       fill = "C-terminal type\n(based on cluster)") +theme_classic(base_size = 22) +  
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, size = 20), axis.text.y = element_text(size = 20), axis.title = element_text(size = 20),legend.title = element_text(size = 16), legend.text = element_text(size = 14))
#  ggsave(path = "Figures", filename = "Tbg.pat_eigen.pdf", device = "pdf", width = 10, height = 6)

df_for_chi_sq_eigen <- dcast(df_eigen,  eigen_type ~ sample) %>%
  mutate_all(funs(replace_na(., 0)))


##do all cross types agree with eigen assignment? 

eigen_test <- All_types_eigen %>%
  filter(sample == "Tb427") %>%
  mutate(agree = ifelse(eigen_type == cross_type, 'yes', 'no') ) %>%
  group_by(agree) 

eigen_test %>% summarize(count = n())
##29 out of 510 VSGs don't match

```

### N-terminal analysis - start here
###Figure 6E
```{r}
#plots

#count of N-type by C-type
CN_df <- read.csv("FigureData/TbgHighlyExpressed.csv") %>%
  select(cluster, nterm_typesubtype, cterm_type) %>%
  distinct() %>%
  group_by(cterm_type, nterm_typesubtype) %>%
  summarise(count = n())
#View(CN_df)


#N-term on x axis
NxC <- read.csv("FigureData/TbgHighlyExpressed.csv") %>%
  select(cluster, nterm_typesubtype, cterm_type) %>%
  distinct() %>%
  group_by(nterm_typesubtype, cterm_type) %>%
  summarise(count = n())
#View(NxC)

NxC %>%
  ggplot(aes(nterm_typesubtype, count, fill = cterm_type)) +
  geom_bar(stat = "identity", width=0.5)  +
  scale_fill_manual(values=color_palette2) +
  #theme(legend.position = "none") +
  labs(title = "", 
       x = "N-terminal subtype", 
       y = "VSG Count",
       fill = "C-terminal type") + theme_classic(base_size = 22) +  
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, size = 20), axis.text.y = element_text(size = 20), axis.title = element_text(size = 20),legend.title = element_text(size = 16), legend.text = element_text(size = 14))
#  ggsave(path = "Figures", filename = "NxC.count.pdf", device = "pdf", width = 10, height = 6)

#proportion
NxC %>%
  ggplot(aes(nterm_typesubtype, count, fill = cterm_type)) +
  geom_bar(position = "fill", stat = "identity", width=0.5)  +
  scale_fill_manual(values=color_palette2) +
  #theme(legend.position = "none") +
  labs(title = "", 
       x = "N-terminal subtype", 
       y = "VSG proportion",
       fill = "C-terminal type") +theme_classic(base_size = 22) +  
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, size = 20), axis.text.y = element_text(size = 20), axis.title = element_text(size = 20),legend.title = element_text(size = 16), legend.text = element_text(size = 14))
#  ggsave(path = "Figures", filename = "NxC.prop.pdf", device = "pdf", width = 10, height = 6)

```


###Figure 6F
```{r}
#C-type by location
loc_C <- read.csv("FigureData/TbgHighlyExpressed.csv") %>%
  select("location", "patient_number", "cluster", "cterm_type") %>%
  distinct() %>%
  group_by(location, cterm_type) %>%
  summarise(count = n())

loc_C %>%
  ggplot(aes(location, count, fill = cterm_type)) +
  geom_bar(stat = "identity", width=0.5)  +
  scale_fill_manual(values=color_palette2) +
  #theme(legend.position = "none") +
  labs(title = "", 
       x = "", 
       y = "VSG count",
       fill = "C-term") + theme_classic(base_size = 22) +  
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, size = 20), axis.text.y = element_text(size = 20), axis.title = element_text(size = 20),legend.title = element_text(size = 16), legend.text = element_text(size = 14))
#  ggsave(path = "Figures", filename = "locC.count.pdf", device = "pdf", width = 12, height = 6)
```

##C terminal statistics
```{r}

All_types_eigen %>% group_by(sample, eigen_type) %>% summarise(n = n())

# make a matrix of counts
contingencyCterm <- matrix(c(172,174,141,23,4,4,7,2,14,28,1,0), nrow = 4, ncol = 3, byrow = F, dimnames = list(c("1", "2-1/2-2", "3", "5"), c("Lister 427", "TbgDAL 972", "tbg patient")))
contingencyCterm

# calculate expected proportions of each type according to genome composition
expected427 <- c(172/510, 174/510, 141/510, 23/510)
expectedDAL <- c(4/17, 4/17, 7/17, 2/17)

# compare patient counts to 427 type frequency
chisq.test(contingencyCterm[,3], p=expected427)

# compare patient counts to DAL972 type frequency
chisq.test(contingencyCterm[,3], p=expectedDAL)

# compare 427 to 972
chisq.test(contingencyCterm[,c(1,2)]) #too few observations for 972 to be confident
# instead look at 427 counts compared to proportions of DAL972 types
chisq.test(contingencyCterm[,1], p=expectedDAL)


# fisher exact test is better for low counts
# full contingency table
fisher.test(contingencyCterm)
# hybrid=T does not make much difference in p-value
fisher.test(contingencyCterm, hybrid=T)
# all >2x2 methods have p-value that meets significance cutoff
fisher.test(contingencyCterm, simulate.p.value = T)

# lister427 v. tbgDAL
fisher.test(contingencyCterm[,c(1,2)])

#lister427 v. tbg patient
fisher.test(contingencyCterm[,c(1,3)])

# tbgDAL v. tbg patient
fisher.test(contingencyCterm[,c(2,3)])

```