---
title: "Supplement-heatmaps"
author: "Jaime So"
date: "8/27/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(reshape2)
library(pheatmap)
```

* the TbgAssembledVSG.csv file includes not the ~1000 total VSG sequences produced by Trinity but all of the ones that met the 1% expression cutoff in at least one library replicate
* TbgHighlyExpressed.csv contains the 44 final VSGs which met the 1% cutoff in at least 2 of 3 tenchical replicates and also includes C-terminal types as determined by network plot

#supplement Fig. 1: T.b. gambiense heatmap
```{r}

tbg_hm <- read_csv("FigureData/TbgAssembledVSG.csv") %>% select(file, Percent, cluster)
tbg_hm_data <- tbg_hm
colnames(tbg_hm_data) <- c("file", "value", "cluster")

tbg_hm_data$cluster <- sub("Cluster", "Gambiense", tbg_hm_data$cluster)

tbg_hm_data_c <- dcast(tbg_hm_data, cluster ~ file, value.var = "value")
tbg_hm_data_c
tbg_hm_data_c2 <- tbg_hm_data_c[2:length(tbg_hm_data_c)]
rownames(tbg_hm_data_c2) <- tbg_hm_data_c$cluster
tbg_hm_m <- as.matrix(tbg_hm_data_c2)
tbg_hm_m <- tbg_hm_m[order(-rowMeans(tbg_hm_m)),]
tbg_hm_m[is.na(tbg_hm_m)] <- 0
tbg_hm_m <- tbg_hm_m + 0.1
tbg_hm_m <- log10(tbg_hm_m)
tbg_hm_m

# color columns by location
tbgannotation.df <- data.frame(unique(read_csv("FigureData/TbgAssembledVSG.csv") %>% .[c(1,8)]))
colnames(tbgannotation.df) <- c("file", "Location")
rownames(tbgannotation.df) <- tbgannotation.df$file[]
tbgannotation.df <- tbgannotation.df %>%
  select(Location)

# color rows by VSG n-terminal subtype
tbgtypes.df <- data.frame(unique(read_csv("FigureData/TbgAssembledVSG.csv") %>% select(cluster, nterm_typesubtype)))
tbgtypes.df$cluster <- sub("Cluster", "Gambiense", tbgtypes.df$cluster)
colnames(tbgtypes.df) <- c("cluster", "Subtype")
rownames(tbgtypes.df) <- tbgtypes.df$cluster[]
tbgtypes.df <- tbgtypes.df %>% select(Subtype)

#palette <- c("skyblue", "steelblue1", "steelblue", "tomato", "red3", "lightgrey")

colors_each <- list(Location = c("A"="darkgoldenrod1",
                                 "B"="olivedrab4",
                                 "C"="darkcyan",
                                 "D"="maroon4",
                                 "E"="orangered2"),
                    Subtype = c("A1"="skyblue",
                                "A2"="steelblue1",
                                "A3"="steelblue",
                                "B1"="tomato",
                                "B2"="red3"))

tbg_heatmap <- pheatmap(tbg_hm_m,
                        breaks = c(seq(-1, -0.1, length.out = 101), 
                                          seq(0, 2, length.out = 101)),
                        fontsize_row = 6,
                        show_rownames = T,
                        cluster_cols = F,
                        annotation_col = tbgannotation.df,
                        annotation_row = tbgtypes.df,
                        cluster_rows = T,
                        treeheight_row = 0,
                        treeheight_col = 0,
                        color = colorRampPalette(c("white", "white", "gray8"))(201),
                        clustering_method = "complete",
                        gaps_col = c(3,6,9,12,15,18,21,24,27,30,33),
                        annotation_colors = colors_each)
tbg_heatmap

```

##Supplement Fig. 5: T.b. rhodesiense
```{r}

# remember to exclude patient 74 from figure
tbr_hm = read_csv("FigureData/Mulindwa2018_rhodesienseVSGtyped.csv") %>% .[!.$patient_number == "74", ]  %>% select(file, Percent, cluster)
tbr_hm_data = tbr_hm
colnames(tbr_hm_data) <- c("file", "value", "cluster")

tbr_hm_data$cluster <- sub("Cluster", "Rhodesiense", tbr_hm_data$cluster)

tbr_hm_data_c <- dcast(tbr_hm_data, cluster ~ file, value.var = "value")
tbr_hm_data_c
tbr_hm_data_c2 <- tbr_hm_data_c[2:length(tbr_hm_data_c)]
rownames(tbr_hm_data_c2) <- tbr_hm_data_c$cluster
tbr_hm_m <- as.matrix(tbr_hm_data_c2)
tbr_hm_m <- tbr_hm_m[order(-rowMeans(tbr_hm_m)),]
tbr_hm_m[is.na(tbr_hm_m)] <- 0
tbr_hm_m <- tbr_hm_m + 0.01
tbr_hm_m <- log10(tbr_hm_m)
tbr_hm_m

# color columns by fluid
tbrannotation.df <- data.frame(unique(read_csv("FigureData/Mulindwa2018_rhodesienseVSGtyped.csv") %>% .[c(1,3)]))
colnames(tbrannotation.df) <- c("file", "Fluid")
rownames(tbrannotation.df) <- tbrannotation.df$file[]
tbrannotation.df <- tbrannotation.df %>%
  select(Fluid)

# color rows by VSG n-terminal subtype
tbrtypes.df <- data.frame(unique(read_csv("FigureData/Mulindwa2018_rhodesienseVSGtyped.csv") %>% select(cluster, nterm_typesubtype)))
tbrtypes.df$cluster <- sub("Cluster", "Rhodesiense", tbrtypes.df$cluster)
colnames(tbrtypes.df) <- c("cluster", "Subtype")
rownames(tbrtypes.df) <- tbrtypes.df$cluster[]
tbrtypes.df <- tbrtypes.df %>% select(Subtype)


#palette <- c("skyblue", "steelblue1", "steelblue", "tomato", "red3", "lightgrey")

colors_each = list(Fluid = c("blood"="maroon4",
                             "CSF"="darkcyan"),
                   Subtype = c("A1"="skyblue",
                               "A2"="steelblue1",
                               "A3"="steelblue",
                               "B1"="tomato",
                               "B2"="red3",
                               "unknown"="lightgrey"))

tbr_heatmap = pheatmap(tbr_hm_m, 
                               breaks = c(seq(-3, -2, length.out = 101), 
                                          seq(-1.9, 2, length.out = 101)),fontsize_row = 6, 
                               show_rownames = T, 
                               cluster_cols = F, 
                               annotation_col = tbrannotation.df, 
                               annotation_row = tbrtypes.df,
                               cluster_rows = T, 
                               treeheight_row = 0, 
                               treeheight_col = 0, 
                               color = colorRampPalette(c("white", "white", "gray8"))(201), 
                               clustering_method = "complete", 
                               annotation_colors = colors_each)
tbr_heatmap

```

