---
title: "N-terminal Network Plot"
author: "Jaime So"
date: "9/2/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(reshape2)
library(readxl)
library(igraph)
```

##Import tables of known N-terminal types which will color code our reference set to assign types to clusters
```{r}

# previous typing data has different VSG gene names, rename them
knownTypes <- read_xlsx("FigureData/All_VSG_BLASTtreeNterm.types.xlsx")
knownTypes <- knownTypes[grep("Tb427", knownTypes$ID), ]
translateID <-read_tsv("FigureData/GAMC_VSGnome2Tb427_2018.translation.txt",col_names = F)
colnames(translateID) <- c("old_name", "new_name")

nameChange <- inner_join(knownTypes[grep("VSG", knownTypes$ID), ], translateID, by = c("ID"="old_name"))
nameChange <- nameChange %>% select(new_name, Nterm)
colnames(nameChange) <- c("ID", "Nterm")

knownTypes <- knownTypes[-grep("VSG", knownTypes$ID), ]
knownTypes <- rbind(knownTypes, nameChange)

knownTypes$colors <- ifelse(knownTypes$Nterm == "A1", "skyblue", knownTypes$Nterm)
knownTypes$colors <- ifelse(knownTypes$Nterm == "A2", "steelblue1", knownTypes$colors)
knownTypes$colors <- ifelse(knownTypes$Nterm == "A3", "steelblue", knownTypes$colors)
knownTypes$colors <- ifelse(knownTypes$Nterm =="B1", "tomato", knownTypes$colors)
knownTypes$colors <- ifelse(knownTypes$Nterm =="B2", "red3", knownTypes$colors)

#import cluster reference table to rename patient VSG by cluster name rather than Trinity ID
cluster_reference_table <- read_csv("FigureData/NetworkPlot_clusterReference.csv")

```

##Import blast output, these will be network links

* the all-vs-all protein blast done to produce the table of links included rhodesiense patient, gambiense patient, and all Tb427 VSG sequences
* this file of pairwise Blast results was too big to upload to github
* since C-terminal network analysis was only done for gambiense patients, the following plots will only include the tb427 reference and highly expressed gambiense VSG
* the file NterminalNetwork.csv includes the filtered pairwise e-values

```{r}

Nlinks <- read_csv("FigureData/NterminalNetwork.csv", col_names=F)
colnames(Nlinks) <- c("from", "to", "evalue")
Nlinks$from <- as.character(Nlinks$from)
Nlinks$to <- as.character(Nlinks$to)

# filter Nlinks so that we include only expressed patient VSGs
# import tables of expressed patient VSG
# rhodesiense data from Christine Clayton's group, hence CC
CCexpressed <- read_csv("FigureData/Mulindwa2018_rhodesienseVSGtyped.csv")
#remove patient 74
CCexpressed <- CCexpressed[!CCexpressed$patient_number==74, ]
CCexpressed$cluster <- sub("Cluster", "Rhodesiense", CCexpressed$cluster)
CCexpressed <- CCexpressed %>% select(cluster, original_seq_length, nterm_seq_length, nterm_typesubtype) %>% .[!duplicated(.$cluster), ]

# gambiense data sequenced by Sarah, so SS
SSexpressed <- read_csv("FigureData/TbgAssembledVSG.csv")
SSexpressed$cluster <- sub("Cluster", "Gambiense", SSexpressed$cluster) 
# filter to include just the highly expressed VSG
SS_list <- SSexpressed %>% dplyr::group_by(patient_number) %>% dplyr::count(cluster) %>% dplyr::filter(n>1) %>% dplyr::ungroup() 
SSexpressed <- SSexpressed[SSexpressed$cluster %in% SS_list$cluster, ] %>% select(cluster, original_seq_length, nterm_seq_length, nterm_typesubtype) %>% distinct()

patient_hmmscan <- bind_rows(SSexpressed, CCexpressed)
colnames(patient_hmmscan)[1] <-"original_seqID"
patient_hmmscan

expressed <- c(SSexpressed$cluster, CCexpressed$cluster)

#add to knownTypes table as unknowns so we can color code them in network plots
unknowns <- tibble(ID = expressed)
unknowns$Nterm <- ifelse(grepl("Gambiense", unknowns$ID), "Gam", "Rho")
unknowns$colors <- ifelse(grepl("Gambiense", unknowns$ID), "white", "yellow")

knownTypes <- bind_rows(knownTypes, unknowns)

# there may be too many nodes to plot, if this happens we will see a cloud rather than distinct clusters
# may want to filter out pseudogenes or fragments
# the hmm profile typing output has sequence length information that will allow me to filter out sequence ID's for fragments. a complete VSG should be ~400 AA's long

#import hmm pipeline output for Lister 427
hmmscan <- read_csv("FigureData/VSG.TbruceiLister427_2018_NtermSummary.csv")
#select relevant columns and filter for most likely complete VSGs >400AA
filter_hmmscan <- hmmscan %>% select(original_seqID, original_seq_length, nterm_seq_length ,nterm_typesubtype) %>% filter(original_seq_length>400 & nterm_seq_length>100) %>% filter(!grepl("pseudogene", .$original_seqID))

#add the hmm types from the patient sets to filter_hmmscan
filter_ref <- bind_rows(filter_hmmscan, patient_hmmscan)
filter_ref <- inner_join(filter_ref, knownTypes, by = c("original_seqID" = "ID"))


#curate node information
# node information will be used later to color code plot based on other typing methods
nodes <- arrange(unique(Nlinks[1]), from)
node_info <- inner_join(nodes, filter_ref, by = c("from" = "original_seqID")) %>% select(from, Nterm, colors)

```


```{r}

# legends for the known VSG plot
typeVSGpalette <- c("skyblue", "steelblue1", "steelblue", "tomato", "red3", "white")
group_ids <- lapply(node_info %>% split(.$Nterm), function(grp) { grp$from })

# here I will only plot reference set and the gambiense patients
Gambiense <- Nlinks[!(grepl("Rhodesiense", Nlinks$from)), ] %>% .[!(grepl("Rhodesiense", .$to)), ]

gambiense_cutoff3 <- filter(Gambiense, evalue < 1e-3)
gambiense3 <- graph_from_data_frame(d=gambiense_cutoff3, vertices=node_info[!(grepl("Rhodesiense", node_info$from)), ], directed=F)
gambiense_simple3 <- igraph::simplify(gambiense3, remove.multiple = T, remove.loops = T)
gambiense_cluster3 <- cluster_leading_eigen(gambiense_simple3)


gambiense_community3 <- gambiense_cluster3$membership
# extract seq ID from the network plot dataframe
gv3 <- as_ids(V(gambiense_simple3))
# each of these can be the column of the dataframe
gv3com <- data.frame(gv3,gambiense_community3)
colnames(gv3com) <- c("original_seqID", "eigen_cluster")
gv3com <- merge(gv3com, filter_ref)
gv3com$eigen_cluster <- as.factor(gv3com$eigen_cluster)

gcom3_ids <- lapply(gv3com %>% split(.$eigen_cluster), function(grp) { grp$VSG })
gcolor3 <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00")
gv3com$Colors <- gcolor3[gv3com$eigen_cluster]

par(mfrow=c(1,2))
set.seed(1234) #keeps the orientation the same for each graph iteration. Don't delete even though it seems redundant
plot(gambiense_cluster3 #cluster communities
     , gambiense_simple3 #maps the vertices and edges
     , edge.color = "grey50" 
     , vertex.size = 7
     , asp = 0
     , edge.arrow.size = 0
     , edge.width = 1
     , vertex.label = NA)
title("leading eigen clusters", cex.main = 1) #plot title
legend("bottomleft", bty = "n",
       legend = levels(gv3com$eigen_cluster),
       col= color3[as.numeric(levels(gv3com$eigen_cluster))],pch = 19, pt.cex = 1.2, cex=0.8, horiz = FALSE)

set.seed(1234)
plot(gambiense_simple3
     , edge.color = "grey50" 
     , vertex.size = 7
     , asp = 0
     , edge.arrow.size = 0
     , edge.width = 2
     , vertex.label = NA
     , vertex.color = node_info[!(grepl("Rhodesiense", node_info$from)), ] %>% .$colors)
title("1e-3 evalue cutoff", cex.main = 1)
legend('bottomleft', legend = names(group_ids[1:6]),
       col = typeVSGpalette[1:6],
       pch = 19, bty = "n",  pt.cex = 1.5, cex = 0.8, 
       text.col = "black", horiz = FALSE)

communities(gambiense_cluster3)

gambiense_cutoff2 <- filter(Gambiense, evalue < 1e-2)
gambiense2 <- graph_from_data_frame(d=gambiense_cutoff2, vertices=node_info[!(grepl("Rhodesiense", node_info$from)), ], directed=F)
gambiense_simple2 <- igraph::simplify(gambiense2, remove.multiple = T, remove.loops = T)
gambiense_cluster2 <- cluster_leading_eigen(gambiense_simple2)


gambiense_community2 <- gambiense_cluster2$membership
# extract seq ID from the network plot dataframe
gv2 <- as_ids(V(gambiense_simple2))
# each of these can be the column of the dataframe
gv2com <- data.frame(gv2,gambiense_community2)
colnames(gv2com) <- c("original_seqID", "eigen_cluster")
gv2com <- merge(gv2com, filter_ref)
gv2com$eigen_cluster <- as.factor(gv2com$eigen_cluster)

gcom2_ids <- lapply(gv2com %>% split(.$eigen_cluster), function(grp) { grp$VSG })
gcolor2 <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00")
gv2com$Colors <- gcolor2[gv2com$eigen_cluster]

par(mfrow=c(1,2))
set.seed(1234) #keeps the orientation the same for each graph iteration. Don't delete even though it seems redundant
plot(gambiense_cluster2 #cluster communities
     , gambiense_simple2 #maps the vertices and edges
     , edge.color = "grey50" 
     , vertex.size = 7
     , asp = 0
     , edge.arrow.size = 0
     , edge.width = 0.8
     , vertex.label = NA)
title("leading eigen clusters", cex.main = 1) #plot title
legend("bottomleft", bty = "n",
       legend = levels(gv2com$eigen_cluster),
       col= gcolor2[as.numeric(levels(gv2com$eigen_cluster))],pch = 19, pt.cex = 1.2, cex=0.8, horiz = FALSE)

set.seed(1234)
plot(gambiense_simple2
     , edge.color = "grey50" 
     , vertex.size = 7
     , asp = 0
     , edge.arrow.size = 0
     , edge.width = 0.8
     , vertex.label = NA
     , vertex.color = node_info[!(grepl("Rhodesiense", node_info$from)), ] %>% .$colors)
title("1e-2 evalue cutoff", cex.main = 1)
legend('bottomleft', legend = names(group_ids[1:6]),
       col = typeVSGpalette[1:6],
       pch = 19, bty = "n",  pt.cex = 1.5, cex = 0.8, 
       text.col = "black", horiz = FALSE)

communities(gambiense_cluster2)

```
